Allocation Removal by Partial Evaluation in a Tracing JIT

### 2011-04-21 talk/icooolps2011/paper.tex
submitted version
Stephen Simmons cannot make it.  But iko can.

\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^a$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany

@@

@@

We already explored promotion in other context, such as earlier versions of
PyPy's JIT as well as a Prolog partial evaluator
\cite{bolz_towards_2009}. Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

>>>>

\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^{a, c}$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany

@@

\terms
Languages, Performance, Experimentation

\keywords{Tracing JIT, Runtime Feedback, Interpreter, Code Generation, Meta-Programming}

@@

We already explored promotion in other context, such as earlier versions of
PyPy's JIT.
%as well as a Prolog partial evaluator \cite{bolz_towards_2009}
Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

### 2011-04-21 talk/pepm2011/presentation/talk.tex
Stephen Simmons cannot make it.  But iko can.

\begin{frame}
  \frametitle{What to do?}
  \begin{itemize}
      \item Use a JIT compiler
      \item \textbf{Add an optimization that can deal with heap operations}
      \pause
      \item optimize short-lived objects
      \item remove some of the redundancy
  \end{itemize}
\end{frame}

@@

\begin{frame}[containsverbatim]
  \frametitle{Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{verbatim}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
x = new(Integer)
set(x, intval, i3)
\end{verbatim}
\begin{verbatim}
guard_class(x, Integer)
guard_class(c, Integer)
i4 = get(x, intval)
i5 = get(c, intval)
i6 = int_add(i4, i5)
y = new(Integer)
set(y, intval, i6)
\end{verbatim}
\end{frame}

\begin{frame}[containsverbatim]
  \frametitle{Optimized Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{alltt}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
\redsout{x = new(Integer)}
\redsout{set(x, intval, i3)}
\end{alltt}
\begin{alltt}
\redsout{guard_class(x, Integer)}
guard_class(c, Integer)
\redsout{i4 = get(x, intval)}
i5 = get(c, intval)
i6 = int_add(\emph{i3}, i5)
y = new(Integer)
set(y, intval, i6)
\end{alltt}
\end{frame}

@@

@@

\begin{frame}
  \frametitle{Benchmark Results}
  \begin{itemize}
      \item to evaluate the optimization we used PyPy's Python interpreter with real-world programs
      \item optimization can remove
      \begin{itemize}
          \item 70\% of all \texttt{new} operations
          \item 14\% of all \texttt{get/set} operations
          \item 93\% of all \texttt{guard} operations
      \pause
      \item Timings improve by a factor between 1.1 and 6.95
      \item outperforming standard Python on all benchmarks but two
      \end{itemize}
  \end{itemize}
\end{frame}

>>>>

\begin{frame}
  \frametitle{What to do?}
  \begin{itemize}
      \item Use a JIT compiler
      \item \textbf{Add an optimization that can deal with heap operations}
      \pause
      \item optimize short-lived objects
      \item remove some of the redundant type checks
  \end{itemize}
\end{frame}

@@

\begin{frame}[containsverbatim]
  \frametitle{Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{verbatim}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
x = new(Integer)
set(x, intval, i3)
\end{verbatim}
\begin{verbatim}
guard_class(x, Integer)
guard_class(c, Integer)
i4 = get(x, intval)
i5 = get(c, intval)
i6 = int_add(i4, i5)
y = new(Integer)
set(y, intval, i6)
return(y)
\end{verbatim}
\end{frame}

\begin{frame}[containsverbatim]
  \frametitle{Optimized Example Trace}
  Trace of \texttt{x = a + b; y = x + c}:
\begin{alltt}
guard_class(a, Integer)
guard_class(b, Integer)
i1 = get(a, intval)
i2 = get(b, intval)
i3 = int_add(i1, i2)
\redsout{x = new(Integer)}
\redsout{set(x, intval, i3)}
\end{alltt}
\begin{alltt}
\redsout{guard_class(x, Integer)}
guard_class(c, Integer)
\redsout{i4 = get(x, intval)}
i5 = get(c, intval)
i6 = int_add(\emph{i3}, i5)
y = new(Integer)
set(y, intval, i6)
return(y)
\end{alltt}
\end{frame}

@@

\begin{frame}
  \frametitle{Properties of the Optimization}
  \begin{itemize}
      \item output trace is never longer than input trace
      \item runtime linear in the length of the trace
  \end{itemize}
  \pause
  \begin{block}{Implementation}
      \begin{itemize}
          \item about 400 lines of code
          \item some added complexity over presentation
          \item objects with arbitrary numbers of fields
          \item array support
      \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Benchmark Results}
  \begin{itemize}
      \item to evaluate the optimization we used PyPy's Python interpreter with real-world programs
      \begin{itemize}
          \item interpreter is about 30'000 lines of code
      \end{itemize}
      \pause
      \item optimization can remove
      \begin{itemize}
          \item 70\% of all \texttt{new} operations
          \item 14\% of all \texttt{get/set} operations
          \item 93\% of all \texttt{guard} operations
      \end{itemize}
      \pause
      \item Timings improve by a factor between 1.1 and 6.95
      \item outperforming standard Python on all benchmarks but two
  \end{itemize}
\end{frame}
